Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-08-06 Paper/assa.tex
..

\label{sub:gsa}
Consider a function $f(\xi_1, \xi_2, \ldots, \xi_\Np)$. 
Whereas the active subspace framework described above not make any assumptions
about statistical independence of the input parameters $\xi_i$, $i = 1, \ldots, \Np$, 
the classical 
framework of variance based sensitivity analysis~\cite{Sobol:2001, Saltelli:2010} assumes the input
parameters are statistically independent. While extensions to the cases 
of correlated inputs exist, we limit the discussion in this section to the
case of random inputs that are statistically independent and are 
either uniformly distributed or
distributed according to a Boltzmann probability distribution.
Note that a measure $\mu$
on $\R$ is referred to as a Boltzmann measure if it is 
absolutely continuous with respect to the Lebesgue measure  
and admits a density  of the form $\pi(x) = C \exp\{-V(x)\}$,
where $V$ is a continuous function and $C$ a normalization 
constant~\cite{Lamboni:2013}.

@@

It was shown in~\cite{Lamboni:2013} that the total Sobol' 
index $T_i(f)$ can be bounded in terms of $\nu_i(f)$:
\begin{equation}\label{equ:sobol_bound}
T_i(f) \leq \frac{C_i}{\V(f)}\nu_i(f), \quad i = 1, \ldots, \Np,
\end{equation}
where for each $i$, $C_i$ is an appropriate \emph{Poincar\'{e}} constant
that depends on the distribution of $\xi_i$.
This provides a strong theoretical basis for using DGSMs to identify 
unimportant inputs. Moreover, 
in~\cite{Vohra:2018} we introduced the normalized screening metric
\[
   \tilde{\nu}_i(f) = \frac{C_i \nu_i(f)}{\sum_{i=1}^\Np C_i \nu_i(f)},
\]
for detecting unimportant input parameters. 
Note that if $\xi_i$ are iid, the $C_i$'s will cancel and the normalized screening
metric reduces to 
\[
    \tilde{\nu}_i(f) = \frac{\nu_i(f)}{\sum_{i=1}^\Np \nu_i(f)} 
      = \frac{\vec{e}_i^T \mat{C} \vec{e}_i}{\trace(\mat{C})} 
      = \frac{\sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2}{\sum_{k = 1}^\Np \lambda_k}.
\]

@@

Combining Proposition~\ref{prp:dgsm_bound} and~\eqref{equ:sobol_bound}, shows
an interesting link between the activity scores and total Sobol' indices.
Specifically, by computing the activity scores, we can identify unimportant
parameters.  
One can attempt to reduce parameter dimension by fixing
unimportant parameters at nominal values. Suppose $\xi_i$ is
deemed unimportant as a result of a global sensitivity analysis
based on computing the activity scores. We want to understand
what is the approximation error that occurs once we fix $\xi_i$ 
at a nominal value.
%Let $\I \subset \{1, \ldots,
%Np\}$ be an index set that indexes the unimportant parameters in the parameter
%vector $\vec{\xi}$.  Letting $|\I|$ denote the number of elements of $\I$, 
%we denote by $\vec{z} \in \R^{|\I|}$ a generic vector of nominal values for
%the unimportant parameters $\vec\xi_\I$, and let $\vec{y}$ be a generic vector of
%important variables, $\vec\xi_{\I^c}$,
%where $\I^c$ denotes the complement of $\I$ in $\{1, \ldots, \Np\}$. 
%For a vector $\xi$, we let $\xin$ denote the vector obtained by removing
%$\xi_i$.

@@

\begin{theorem}
We have $\Ez{ \delta(z)} \leq 2C_i\big(\act{f}{i}{r} + \lambda_{r+1}\big)/{\V(f)} $.
\end{theorem}
%
%\[
%       T_i(f) \leq \frac{C_i}{\V(f)}(\act{f}{i}{r} + \lambda_{r+1}).
%\]
%we obtain the following useful inequality, which
%links activity scores and Sobol' indices:
%for $1 \leq r \leq \Np$, and $i = 1, \ldots, \Np$, we have
%\[
%       T_i(f) \leq \frac{C_i}{\V(f)}(\act{f}{i}{r} + \lambda_{r+1}).
%\]
%This was noted in~\cite{Diaz:2016} for the specific case of all $\xi_i$'s being
%iid $U(0, 1)$, in which case $C_i = 1/\pi^2$, $i = 1, \ldots, \Np$.
\begin{proof} \hayleynote{(NOT FINISHED YET, NEED TO SORT OUT INDEXING TO BE IN LINE WITH THE PAPER)}
First we show the numerator 
\[\int_\Omega \big( f(\vec\xi) - f^{z}(\vec\xi)\big)^2 \, \mu(d\vec\xi) = 2\V_{U^c}(f) \]
Using the ANOVA decomposition we write:
\[ f(\vec\xi) = f_0 + f_1(\vec\xi_{U}) + f_2(\vec\xi_{U^C}) + f_{12}(\vec\xi_U,\vec\xi_{U^C})\]
Then we can simplify the numerator as follows:
\[ \int_{\Omega} \big( f_2(\vec\xi_{U^C}) + f_{12}(\vec\xi_{U},\vec\xi_{U^C}) - f_2(z) -f_{12}(\vec\xi_{U},z)\big)^2\mu(d\vec\xi) \]
Using the properties of ANOVA:
\[
\int_{\Omega_{U^c}} f_2(\vec\xi_{{U^C}}) \,\mu(d\vec\xi_{U^C}) 
= \int_{\Omega_{U^C}} f_{12}(\vec\xi_U, \vec\xi_{U^C})\,\mu(d\vec\xi_{U^C}) 
= \int_{\Omega_U} f_{12}(\vec\xi_U, \vec\xi_{U^C})\,\mu(d\vec\xi_U) = 0,
\]
For a fixed $z \in \Omega_{U^C}$ we can simplify the numerator further:
\[ \int_{\Omega}  f_2^2(\vec\xi_{U^C}) + f_{12}^2(\vec\xi_{U},\vec\xi_{U^C}) - f_2^2(z) -f_{12}^2(\vec\xi_{U},z) \mu(d\vec\xi) = \]
\[ \V_{U^C}(f) + \V_{U^C,U}(f) + f_2^2(z) + \int_{\Omega_U} f_{12}^2(\vec\xi_U, z) \mu(d\vec\xi_U),\]
Taking the expectation of the above expression over $\Omega_{U^c}$ gives 
desired result:
\[\V_{U^C}(f) + \V_{U^C,U}(f) + \int_{\Omega_{U^C}} f_2^2(z) \, \mu(dz)
+ \int_{\Omega_{U^C}}\int_{\Omega_U} f_{12}^2(\vec\xi_U,z)\,\mu (d\vec\xi_U)
\mu (dz) \]
\[= \V_{U^C}(f) + \V_{U,U^C}(f) + \V_{U^C}(f) + \V_{U,U^c}(f) 
= 2 \V_{U^c}^\text{tot}(f) \]
Using the definition of variance we rewrite the the denominator as:
\[
 \int_\Omega f(\vec\xi)^2\, \mu(d\vec\xi) 
=  \V(f) + 
\Big(\int_\Omega f(\vec\xi)\mu(d\vec\xi)\Big)^2 
\geq   \V(f) 
\]	
Combining by parts we have:
\[ \Ez{ \delta(z)} = 2T_{U^C}(f) \]
Using (10) above yields:
\[2T_{U^C}(f) \leq \frac{C_{U^C}}{\V(f)}\nu_{U^C}(f) \]
Finally using Proposition 2.1 gives us the result:
\[2T_{U^C}(f) \leq 2\frac{C_{U^C}}{\V(f)}\nu_{U^C}(f) \leq 2\frac{C_{U^C}}{\V(f)}(\nu_{{U^C},r}(f) + \lambda_{r+1} ) \]
\end{proof}

>>>>

\label{sub:gsa}
Consider a function: $f(\xi_1, \xi_2, \ldots, \xi_\Np)$. 
While the active subspace framework described above does not make any assumptions
about statistical independence of the inputs $\xi_i$, $i = 1, \ldots, \Np$, 
the classical 
framework of variance based sensitivity analysis~\cite{Sobol:2001, Saltelli:2010} 
assumes that the inputs
are statistically independent. While extensions to the cases 
of correlated inputs exist~\cite{Borgonovo:2007,Li:2010,Jacques:2006,Xu:2007},
 we limit the discussion in this section to the
case of random inputs that are statistically independent and are 
either uniformly distributed or
distributed according to a Boltzmann probability distribution.
Note that a measure $\mu$
on $\R$ is referred to as a Boltzmann measure if it is 
absolutely continuous with respect to the Lebesgue measure  
and admits a density  of the form $\pi(x) = C \exp\{-V(x)\}$,
where $V$ is a continuous function and $C$ a normalization 
constant~\cite{Lamboni:2013}.

@@

It was shown in~\cite{Lamboni:2013} that the total Sobol' 
index $T_i(f)$ can be bounded in terms of $\nu_i(f)$:
\begin{equation}\label{equ:sobol_bound}
T_i(f) \leq \frac{C_i}{\V(f)}\nu_i(f), \quad i = 1, \ldots, \Np,
\end{equation}
where for each $i$, $C_i$ is an appropriate \emph{Poincar\'{e}} constant
that depends on the distribution of $\xi_i$.
This provides a strong theoretical basis for using DGSMs to identify 
unimportant inputs. Moreover, 
in~\cite{Vohra:2018} we introduced the normalized screening metric
\[
   \tilde{\nu}_i(f) = \frac{C_i \nu_i(f)}{\sum_{i=1}^\Np C_i \nu_i(f)},
\]
for detecting unimportant inputs. 
Note that if $\xi_i$ are iid, the $C_i$'s will cancel and the normalized screening
metric reduces to 
\[
    \tilde{\nu}_i(f) = \frac{\nu_i(f)}{\sum_{i=1}^\Np \nu_i(f)} 
      = \frac{\vec{e}_i^T \mat{C} \vec{e}_i}{\trace(\mat{C})} 
      = \frac{\sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2}{\sum_{k = 1}^\Np \lambda_k}.
\]

@@

Combining Proposition~\ref{prp:dgsm_bound} and~\eqref{equ:sobol_bound}, shows
an interesting link between the activity scores and total Sobol' indices.
Specifically, by computing the activity scores, we can identify the unimportant
inputs.  
One can attempt to reduce parameter dimension by fixing
unimportant inputs at nominal values. Suppose $\xi_i$ is
deemed unimportant as a result of a global sensitivity analysis
based on the activity scores; we want to estimate
the approximation error that occurs once $\xi_i$ 
is fixed at its nominal value.
%Let $\I \subset \{1, \ldots,
%Np\}$ be an index set that indexes the unimportant parameters in the parameter
%vector $\vec{\xi}$.  Letting $|\I|$ denote the number of elements of $\I$, 
%we denote by $\vec{z} \in \R^{|\I|}$ a generic vector of nominal values for
%the unimportant parameters $\vec\xi_\I$, and let $\vec{y}$ be a generic vector of
%important variables, $\vec\xi_{\I^c}$,
%where $\I^c$ denotes the complement of $\I$ in $\{1, \ldots, \Np\}$. 
%For a vector $\xi$, we let $\xin$ denote the vector obtained by removing
%$\xi_i$.

@@

\begin{theorem}
We have $\Ez{ \delta(z)} \leq 2C_i\big(\act{f}{i}{r} + \lambda_{r+1}\big)/{\V(f)} $.
\end{theorem}
%
%\[
%       T_i(f) \leq \frac{C_i}{\V(f)}(\act{f}{i}{r} + \lambda_{r+1}).
%\]
%we obtain the following useful inequality, which
%links activity scores and Sobol' indices:
%for $1 \leq r \leq \Np$, and $i = 1, \ldots, \Np$, we have
%\[
%       T_i(f) \leq \frac{C_i}{\V(f)}(\act{f}{i}{r} + \lambda_{r+1}).
%\]
%This was noted in~\cite{Diaz:2016} for the specific case of all $\xi_i$'s being
%iid $U(0, 1)$, in which case $C_i = 1/\pi^2$, $i = 1, \ldots, \Np$.

### 2018-08-06 Paper/conc.tex
..

Two computational strategies were explored to compute the active subspace for the H$_2$/O$_2$
kinetics application in this work. The first strategy referred to as gradient-based requires computation of gradient of the 
model output with respect to the input, whereas, the second strategy, referred to as gradient-free implements a local
linear approximation to the model output in order to estimate the gradient. The two strategies are shown to yield
consistent results for the 19-dimensional problem whereby only the pre-exponents are considered to be uncertain.
Additionally, the activity scores are also shown to be consistent with the screening metric estimates based on
DGSMs in~\cite{Vohra:2018}. Note that an iterative procedure is adopted for implementing the two strategies
for enhancing computational efficiency. 
Motivated by these findings, the gradient-free approach is used for computing the active subspace in the high-dimensional
case involving 33 uncertain parameters. Based on the prior marginals of the uncertain parameters, our results
indicate that the uncertainty in the ignition delay is largely dependent on $A_1$, $A_9$, and $E_{a,1}$, although
 contributions from $A_{15}$ and $A_{17}$ are also found to be significant.

>>>>

Two computational strategies were explored to compute the active subspace for the H$_2$/O$_2$
kinetics application in this work. The first strategy referred to as gradient-based requires computation of gradient of the 
model output with respect to the input, whereas, the second strategy, referred to as gradient-free implements a local
linear approximation to the model output in order to estimate the gradient. The two strategies are shown to yield
consistent results for the 19-dimensional problem whereby only the pre-exponents are considered to be uncertain.
Additionally, the activity scores are also shown to be consistent with the screening metric estimates based on
DGSMs in~\cite{Vohra:2018}. Note that an iterative procedure is adopted for implementing the two strategies
for enhancing computational efficiency. 
Motivated by these findings, the gradient-free approach is used for computing the active subspace in the high-dimensional
case involving 33 uncertain inputs. Based on the prior marginals of the uncertain inputs, our results
indicate that the uncertainty in the ignition delay is largely dependent on $A_1$, $A_9$, and $E_{a,1}$, although
 contributions from $A_{15}$ and $A_{17}$ are also found to be significant.

### 2018-08-06 Paper/method.tex
..

We begin by evaluating the gradient of the model output, $\nabla_{\bm{\xi}}f$,
at an initial set of $n_0$ samples denoted by $\bm{\xi}_i$, $i$ = $1,\ldots,n_0$.
Using the gradient
evaluations, the matrix, $\hat{\mat{C}}$ is computed. Eigenvalue decomposition
of $\hat{\mat{C}}$ yields an initial estimate of the dominant eigenspace,
$\hat{\mat{W}}_1$ and a diagonal matrix, 
$\hat{\mat{\Lambda}}_1$ that comprises the eigenvalues along the diagonal.
 At each subsequent iteration, model evaluations are
generated at a new set of $n_k$ samples. The new set of gradient evaluations
are augmented with the available set to re-construct $\hat{\mat{C}}$ followed
by its eigenvalue decomposition. The relative change in the norm of the
difference in squared value of individual components of the dominant
eigenvectors between subsequent iterations is evaluated. The process is
terminated and the resulting eigenspace is considered to be converged once
the maximum relative change at iteration $k$, $\delta \hat{\mat{W}}_{1,j}^{(k)}$
($j$ is used as an index for the eigenvectors),
is smaller than a given tolerance, $\tau$.  A regression fit to
$G(\hat{\mat{W}}_1^\top\bm{\xi})$ is used as a surrogate to characterize and
quantify the uncertainty in the model output. Moreover, the components of the
dominant eigenvectors are used to compute the activity scores, $\bm{\nu}_r(f)$,
which provide insight into the relative importance of the uncertain inputs.
Note that the index, $r$, corresponds to the number of dominant eigenvectors in
$\hat{\mat{W}}_1$. The sequence of steps for the gradient-based approach are
provided in Algorithm~\ref{alg:grad}.

@@

We complement the comparative assessment of the two approaches by estimating
the activity scores for individual uncertain parameters ($\nu_{i,r}$; $r$=1 since the
active subspace is 1-dimensional)
using the components of the dominant eigenvector as shown in Algorithm~\ref{alg:grad}.
The activity scores for the 19 uncertain pre-exponents ($A_i$'s), estimated
using the gradient-based and gradient-free approaches are plotted in
Figure~\ref{fig:comp_as} (left).  Additionally, for the purpose of
verification, we compare these estimates with those obtained using the
corresponding DGSMs, as reported in~\cite{Vohra:2018}, in
Figure~\ref{fig:comp_as} (right).

>>>>

We begin by evaluating the gradient of the model output, $\nabla_{\bm{\xi}}f$,
at an initial set of $n_0$ samples denoted by $\bm{\xi}_i$, $i$ = $1,\ldots,n_0$.
Using the gradient
evaluations, the matrix, $\hat{\mat{C}}$ is computed. Eigenvalue decomposition
of $\hat{\mat{C}}$ yields an initial estimate of the dominant eigenspace,
$\hat{\mat{W}}_1$ and the set of corresponding eigenvalues, 
$\hat{\mat{\Lambda}}_1$.
 At each subsequent iteration, model evaluations are
generated at a new set of $n_k$ samples. The new set of gradient evaluations
are augmented with the available set to re-construct $\hat{\mat{C}}$ followed
by its eigenvalue decomposition. The relative change in the norm of the
difference in squared value of individual components of the dominant
eigenvectors between subsequent iterations is evaluated. The process is
terminated and the resulting eigenspace is considered to be converged once
the maximum relative change at iteration $k$, $\delta \hat{\mat{W}}_{1,j}^{(k)}$
($j$ is used as an index for the eigenvectors),
is smaller than a given tolerance, $\tau$.  A regression fit to
$G(\hat{\mat{W}}_1^\top\bm{\xi})$ is used as a surrogate to characterize and
quantify the uncertainty in the model output. Moreover, the components of the
dominant eigenvectors are used to compute the activity scores, $\bm{\nu}_r(f)$,
which provide insight into the relative importance of the uncertain inputs.
Note that the index, $r$, corresponds to the number of dominant eigenvectors in
$\hat{\mat{W}}_1$. The sequence of steps for the gradient-based approach are
provided in Algorithm~\ref{alg:grad}.

@@

We complement the comparative assessment of the two approaches by estimating
the activity scores for individual uncertain inputs ($\nu_{i,r}$; $r$=1 since the
active subspace is 1-dimensional)
using the components of the dominant eigenvector as shown in Algorithm~\ref{alg:grad}.
The activity scores for the 19 uncertain pre-exponents ($A_i$'s), estimated
using the gradient-based and gradient-free approaches are plotted in
Figure~\ref{fig:comp_as} (left).  Additionally, for the purpose of
verification, we compare these estimates with those obtained using the
corresponding DGSMs, as reported in~\cite{Vohra:2018}, in
Figure~\ref{fig:comp_as} (right).
